---
title: "nba injury"
output: pdf_document
date: "2025-06-02"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)    # For data manipulation and visualization
library(lubridate)    # For date handling
library(plotly)       # For interactive plots
library(corrplot)     # For correlation plots
library(skimr)        # For data summary
library(GGally)       # For pair plots
library(scales)
library(dplyr)
library(ggplot2)
library(readr)
library(stringr)
library(DT)
```


```{r cars}
# NBA Comprehensive Injury Analysis - Complete EDA with Location Details
# Updated: 2025-06-06
# Author: ol0w0lo

# --- SETUP AND LIBRARIES ---
library(tidyverse)
library(lubridate)
library(skimr)
library(corrplot)
library(scales)
library(ggrepel)
library(tidytext)
library(forcats)
library(dplyr)
library(gridExtra)
```



```{r}

#bios <- read_csv("~/Downloads/nba_bios_2013-2023.csv")
run <- read_csv("C:/Users/11578/Desktop/running distance nba.csv")
bios <- read_csv("C:/Users/11578/Desktop/nba_bios_2013-2023.csv")
all_injury <- read.csv("C:/Users/11578/Desktop/NBA Player Injury Stats(1951 - 2023).csv", stringsAsFactors = FALSE)
# Basic dataset overview
cat("=== NBA Player Bios Analysis ===\n")
cat("Dataset shape:", nrow(bios), "rows x", ncol(bios), "columns\n")
cat("Seasons covered:", paste(sort(unique(bios$SEASON)), collapse = ", "), "\n")
cat("Number of unique players:", length(unique(bios$PLAYER_NAME)), "\n")

# Clean the dataset remove columns not important
bios <- bios %>% 
  select(-TEAM_ID, -COLLEGE, -COUNTRY, 
         -DRAFT_YEAR, -DRAFT_ROUND, -DRAFT_NUMBER, -PLAYER_HEIGHT)

# Basic statistical summary
cat("\n=== Statistical Summary ===\n")
summary(bios)
head(bios)

```





Teams often put players on IL without them getting injure because it gives the players more time to rest. so in the following code it will exclude the players on 

```{r}
# Load and clean injury data
#all_injury <- read_csv("~/Downloads/NBA Player Injury Stats(1951 - 2023).csv")

injury <- all_injury %>%
  mutate(
    Notes = as.character(Notes),
    # Extract injury description after 'placed on IL with ...'
    injury_type = ifelse(
      str_detect(Notes, regex("placed on IL with ", ignore_case = TRUE)),
      str_trim(str_remove(Notes, regex(".*placed on IL with ", ignore_case = TRUE))),
      NA_character_
    )
  ) %>%
  # Exclude: 
  # - Notes say 'activated from IL'
  # - or 'placed on IL' but NOT 'with ...'
  filter(
    !str_detect(Notes, regex("activated from IL", ignore_case = TRUE)),
    !(str_detect(Notes, regex("placed on IL", ignore_case = TRUE)) &
      !str_detect(Notes, regex("placed on IL with ", ignore_case = TRUE)))
  ) %>%
  # Keep only rows with a real injury description
  filter(!is.na(injury_type)) %>%
  # Remove the Notes column
  select(-Notes) %>%
  # Add cleaned player names and date processing
  mutate(
    PLAYER_NAME_CLEAN = str_trim(toupper(Relinquished)),
    Date = as.Date(Date)
  )

# Extract season from date
injury <- injury %>%
  mutate(
    Season_Year = year(Date),
    SEASON = paste0(Season_Year, "-", str_sub(Season_Year + 1, -2))
  )

# Clean bios for merge
bios <- bios %>%
  mutate(PLAYER_NAME_CLEAN = str_trim(toupper(PLAYER_NAME)))

# Add binary 'injured' flag per player-season
injury_summary <- injury %>%
  group_by(PLAYER_NAME_CLEAN, SEASON) %>%
  summarise(
    injured = TRUE,
    most_common_injury = names(sort(table(injury_type), decreasing = TRUE))[1],
    total_injuries = n(), # Number of times injured that season
    .groups = "drop"
  )

# Merge injury info into bios
bioinjury <- bios %>%
  left_join(injury_summary, by = c("PLAYER_NAME_CLEAN", "SEASON")) %>%
  mutate(
    injured = ifelse(is.na(injured), FALSE, injured),
    most_common_injury = ifelse(is.na(most_common_injury), "None", most_common_injury),
    total_injuries = ifelse(is.na(total_injuries), 0, total_injuries)
  ) %>%
  select(-PLAYER_NAME_CLEAN)

# View the injury filtering result
cat("=== Injury Data After Filtering ===\n")
cat("Total injury records:", nrow(injury), "\n")
cat("Unique players with injuries:", length(unique(injury$PLAYER_NAME_CLEAN)), "\n")

# Preview injury data
injury %>% 
  select(Date, Team, Relinquished, injury_type) %>% 
  head(20)

# Display final merged result
cat("\n=== Final Merged Data ===\n")
head(bioinjury)

# Summary of injury status
cat("\n=== Injury Summary ===\n")
injury_status <- bioinjury %>%
  summarise(
    total_player_seasons = n(),
    injured_player_seasons = sum(injured),
    injury_rate = mean(injured) * 100
  )

cat("Total player-seasons:", injury_status$total_player_seasons, "\n")
cat("Player-seasons with injuries:", injury_status$injured_player_seasons, "\n")
cat("Injury rate:", round(injury_status$injury_rate, 2), "%\n")

```





```{r}
#run <- read_csv("~/Downloads/running distance nba.csv")


run <- run %>%
  mutate(
    PLAYER_NAME_CLEAN = str_trim(toupper(PLAYER_NAME)),
    SEASON = as.numeric(substr(SEASON, 1, 4))  # Ensure SEASON is numeric
  ) %>%
  select(
    PLAYER_NAME_CLEAN, 
    SEASON,
    PLAYER_ID,
    TEAM_ABBREVIATION,
    GP,
    MIN,
    DIST_MILES,
    DIST_MILES_OFF,
    DIST_MILES_DEF,
    AVG_SPEED,
    AVG_SPEED_OFF,
    AVG_SPEED_DEF,
    SEASON_TYPE
  )

# Ensure consistent formatting in bioinjury
bioinjury <- bioinjury %>%
  mutate(
    PLAYER_NAME_CLEAN = str_trim(toupper(PLAYER_NAME)),
    SEASON = as.numeric(substr(SEASON, 1, 4))  # Convert "2018-19" → 2018
  )

# Prepare injury_type_clean with SEASON as numeric
injury_type_clean <- injury %>%
  mutate(
    PLAYER_NAME_CLEAN = str_trim(toupper(Relinquished)),
    SEASON = as.numeric(year(as.Date(Date)))
  ) %>%
  select(PLAYER_NAME_CLEAN, SEASON, injury_type)

# Merge bios + injury status + running data + injury type
merged <- bioinjury %>%
  left_join(
    run, 
    by = c("PLAYER_NAME_CLEAN", "SEASON"),
    suffix = c("", "_run")
  ) %>%
  left_join(
    injury_type_clean,
    by = c("PLAYER_NAME_CLEAN", "SEASON")
  ) %>%
  select(
    -PLAYER_NAME_CLEAN,
    -PLAYER_ID_run,
    -TEAM_ABBREVIATION_run,
    -GP_run
  )

# Display results
cat("=== Merge Complete ===\n")
cat("Total rows:", nrow(merged), "\n")
cat("Players with running data:", sum(!is.na(merged$GP)), "\n")

# Preview merged data with injury type
print(head(merged))

# Filter injured players only
injured_data <- merged %>% 
  filter(injured == TRUE)

# Preview injured data
head(injured_data)
```

```{r}

injurybyyear <- all_injury %>%
  mutate(
    Date = as.Date(Date),
    PLAYER_NAME = str_trim(toupper(Relinquished)),
  ) %>%
  filter(
    !is.na(PLAYER_NAME),
    PLAYER_NAME != "",
    grepl("IL|injury|placed on inactive|out|injured", Notes, ignore.case = TRUE),
    !grepl("illness|stomach|health and safety protocols|COVID|virus|flu|sick|personal|family|bereavement|conditioning|rest|load management", Notes, ignore.case = TRUE)
  ) %>%
  select(Date, Team, PLAYER_NAME, Notes)

# Extract year and create summary
injury_trend_year <- injurybyyear %>%
  mutate(
    Year = year(Date)
  ) %>%
  group_by(Year) %>%
  summarise(
    total_injuries = n(),
    unique_players = n_distinct(PLAYER_NAME),
    .groups = "drop"
  ) %>%
  arrange(Year)

injury_trend_year %>%
  ggplot(aes(x = Year, y = total_injuries)) +
  geom_line(color = "#1f77b4", size = 1.2) +
  geom_point(color = "#ff7f0e", size = 3) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.3, color = "#2ca02c") +
  scale_x_continuous(breaks = seq(from = min(injury_trend_year$Year), 
                                  to = max(injury_trend_year$Year), 
                                  by = 5)) +
  scale_y_continuous(labels = comma_format()) +
  labs(
    title = "NBA Injury Trend Over Time",
    subtitle = "Total number of injury-related IL placements by year",
    x = "Year",
    y = "Total Injuries",
    caption = paste("Data: NBA Injury Analysis")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray60"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )

```
This plot shows the trend of injury in the NBA over the years. There are large variance in the beginning of 1950s and 2022-23. The variance in the early stages of NBA is likely due to the lack of tracking as shown in the graph. And the huge dip in the end of the curve is because this data set only captures till April of 2023, so there are injuries snot counted. An interesting feature shown in the plot shows a decrease in 2020, this was likely because due to COVID, and it cause less games to be played. In that season, the NBA officials reduce the number of game, and ended the season in October, as compared to mid-June normally after playoff. The season after the lock down the injury shown a large increase because there's less time for the players to rest before the new season. 

```{r}
cat("=== Checking Column Names ===\n")
print(colnames(merged))
cat("\n")

# Let's also check the structure
str(merged)

```



```{r}
# NBA Injury Analysis - Unsupervised Machine Learning Models
# Updated for actual data structure
# Author: ol0w0lo
# Date: 2025-06-10

# Load additional libraries for unsupervised learning
library(cluster)        # For clustering algorithms
library(factoextra)     # For cluster visualization
library(NbClust)        # For optimal cluster number determination
library(pheatmap)       # For heatmaps
library(VIM)            # For missing value visualization
library(mice)           # For missing value imputation
library(umap)           # For UMAP dimensionality reduction
library(dbscan)         # For DBSCAN clustering
library(kohonen)        # For Self-Organizing Maps
library(corrplot)       # For correlation analysis
library(plotly)         # For interactive plots
library(gridExtra)      # For arranging plots

# === DATA PREPARATION ===
cat("=== Starting Unsupervised ML Analysis ===\n")
cat("Original data shape:", nrow(merged), "x", ncol(merged), "\n")

# Clean and prepare data for unsupervised learning
ml_data <- merged %>%
  # Remove rows with missing key variables
  filter(!is.na(PLAYER_NAME), !is.na(SEASON), !is.na(AGE)) %>%
  # Select relevant variables for analysis
  select(
    PLAYER_NAME, SEASON, injured, total_injuries, most_common_injury,
    AGE, PLAYER_HEIGHT_INCHES, PLAYER_WEIGHT, 
    GP, PTS, REB, AST, MIN,
    NET_RATING, OREB_PCT, DREB_PCT, USG_PCT, TS_PCT, AST_PCT,
    DIST_MILES, DIST_MILES_OFF, DIST_MILES_DEF,
    AVG_SPEED, AVG_SPEED_OFF, AVG_SPEED_DEF
  ) %>%
  # Remove rows with too many missing values in key performance metrics
  filter(
    !is.na(GP), !is.na(MIN), 
    !is.na(DIST_MILES) | !is.na(PTS)  # At least one performance metric
  )

cat("Cleaned data shape:", nrow(ml_data), "x", ncol(ml_data), "\n")

# === FEATURE ENGINEERING ===
cat("\n=== Feature Engineering ===\n")

# Create additional meaningful features
ml_features <- ml_data %>%
  mutate(
    # Injury risk categories
    injury_risk = case_when(
      total_injuries == 0 ~ "No_Injury",
      total_injuries == 1 ~ "Low_Risk",
      total_injuries <= 3 ~ "Medium_Risk",
      TRUE ~ "High_Risk"
    ),
    
    # Age categories
    age_group = case_when(
      AGE < 23 ~ "Young",
      AGE < 27 ~ "Prime",
      AGE < 31 ~ "Veteran",
      TRUE ~ "Old"
    ),
    
    # BMI calculation
    BMI = (PLAYER_WEIGHT * 703) / (PLAYER_HEIGHT_INCHES^2),
    
    # Performance efficiency metrics
    minutes_per_game = ifelse(GP > 0, MIN / GP, 0),
    points_per_minute = ifelse(MIN > 0, PTS / MIN, 0),
    rebounds_per_minute = ifelse(MIN > 0, REB / MIN, 0),
    assists_per_minute = ifelse(MIN > 0, AST / MIN, 0),
    
    # Movement efficiency (only if movement data available)
    distance_per_minute = ifelse(!is.na(DIST_MILES) & MIN > 0, DIST_MILES / MIN, NA),
    off_def_speed_ratio = ifelse(!is.na(AVG_SPEED_OFF) & !is.na(AVG_SPEED_DEF) & AVG_SPEED_DEF > 0, 
                                 AVG_SPEED_OFF / AVG_SPEED_DEF, NA),
    off_def_distance_ratio = ifelse(!is.na(DIST_MILES_OFF) & !is.na(DIST_MILES_DEF) & DIST_MILES_DEF > 0,
                                    DIST_MILES_OFF / DIST_MILES_DEF, NA),
    
    # Playing style indicators
    usage_efficiency = ifelse(USG_PCT > 0, TS_PCT / USG_PCT, NA),
    rebounding_rate = OREB_PCT + DREB_PCT,
    
    # Season experience (proxy)
    season_numeric = as.numeric(SEASON)
  ) %>%
  # Remove infinite values
  mutate(across(where(is.numeric), ~ifelse(is.infinite(.), NA, .)))

# Select numeric features for clustering (focusing on available data)
numeric_cols <- ml_features %>%
  select(
    AGE, PLAYER_HEIGHT_INCHES, PLAYER_WEIGHT, BMI,
    GP, minutes_per_game, PTS, REB, AST,
    points_per_minute, rebounds_per_minute, assists_per_minute,
    NET_RATING, OREB_PCT, DREB_PCT, USG_PCT, TS_PCT, AST_PCT,
    rebounding_rate, total_injuries
  ) %>%
  # Add movement data if available
  {if("DIST_MILES" %in% names(ml_features) && sum(!is.na(ml_features$DIST_MILES)) > 100) {
    bind_cols(., select(ml_features, DIST_MILES, AVG_SPEED))
  } else .} %>%
  # Keep only complete cases for initial analysis
  drop_na()

# Get corresponding complete cases from ml_features
complete_indices <- which(complete.cases(
  ml_features[, names(numeric_cols)]
))
ml_complete <- ml_features[complete_indices, ]
numeric_features <- as.matrix(numeric_cols)

cat("Features selected for analysis:", ncol(numeric_features), "\n")
cat("Complete cases for analysis:", nrow(numeric_features), "\n")
cat("Feature names:", paste(colnames(numeric_features), collapse = ", "), "\n")

# Scale the features
scaled_features <- scale(numeric_features)

# === 1. EXPLORATORY DATA ANALYSIS ===
cat("\n=== Exploratory Data Analysis ===\n")

# Injury distribution
injury_dist <- table(ml_complete$injury_risk)
cat("Injury risk distribution:\n")
print(injury_dist)

# Age vs Injury relationship
age_injury_plot <- ggplot(ml_complete, aes(x = age_group, fill = injury_risk)) +
  geom_bar(position = "fill") +
  scale_fill_viridis_d() +
  labs(title = "Injury Risk by Age Group", 
       x = "Age Group", y = "Proportion", fill = "Injury Risk") +
  theme_minimal()
print(age_injury_plot)

# === 2. PRINCIPAL COMPONENT ANALYSIS (PCA) ===
cat("\n=== Principal Component Analysis ===\n")

# Perform PCA
pca_result <- prcomp(scaled_features, center = FALSE, scale. = FALSE)

# PCA summary
pca_summary <- summary(pca_result)
cat("PCA Variance Explained:\n")
print(pca_summary$importance[1:3, 1:min(10, ncol(pca_summary$importance))])

pca_3d_data <- data.frame(
  PC1 = pca_result$x[,1],
  PC2 = pca_result$x[,2], 
  PC3 = pca_result$x[,3],
  injury_risk = ml_complete$injury_risk,
  player_name = ml_complete$PLAYER_NAME
)

pca_3d_plot <- plot_ly(pca_3d_data, 
                       x = ~PC1, y = ~PC2, z = ~PC3,
                       color = ~injury_risk,
                       colors = viridis::viridis(4),
                       text = ~player_name,
                       type = "scatter3d", 
                       mode = "markers") %>%
  layout(title = "3D PCA - NBA Players by Injury Risk",
         scene = list(xaxis = list(title = paste("PC1 (", round(pca_summary$importance[2,1]*100, 1), "%)")),
                     yaxis = list(title = paste("PC2 (", round(pca_summary$importance[2,2]*100, 1), "%)")),
                     zaxis = list(title = paste("PC3 (", round(pca_summary$importance[2,3]*100, 1), "%)"))))

pca_3d_plot

# Multiple 2D projections
pc_combinations <- list(
  c(1,2), c(1,3), c(1,4), c(2,3), c(2,4), c(3,4)
)

create_pca_plot <- function(pc_pair, data, pca_summary) {
  pc_data <- data.frame(
    PC_X = pca_result$x[,pc_pair[1]],
    PC_Y = pca_result$x[,pc_pair[2]],
    injury_risk = ml_complete$injury_risk
  )
  
  var_x <- round(pca_summary$importance[2,pc_pair[1]]*100, 1)
  var_y <- round(pca_summary$importance[2,pc_pair[2]]*100, 1)
  
  ggplot(pc_data, aes(x = PC_X, y = PC_Y, color = injury_risk)) +
    geom_point(alpha = 0.6, size = 1.5) +
    scale_color_viridis_d() +
    labs(title = paste("PC", pc_pair[1], "vs PC", pc_pair[2]),
         x = paste("PC", pc_pair[1], "(", var_x, "%)"),
         y = paste("PC", pc_pair[2], "(", var_y, "%)"),
         color = "Injury Risk") +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# Create multiple PC plots
pc_plots <- map(pc_combinations[1:4], ~create_pca_plot(.x, pca_result, pca_summary))
grid.arrange(grobs = pc_plots, ncol = 2)

# PCA Variables plot
pca_var_plot <- fviz_pca_var(pca_result,
                             col.var = "contrib",
                             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                             title = "PCA - Variable Contributions") +
  theme_minimal()
print(pca_var_plot)

# Scree plot
scree_plot <- fviz_eig(pca_result, addlabels = TRUE, ylim = c(0, 50)) +
  labs(title = "PCA Scree Plot - Variance Explained") +
  theme_minimal()
print(scree_plot)

```
Injury risk by age group:
We categorized players into 4 categories according to age:Young, Veteran, Prime and old; We categorize the risk by High-risk, Low-risk, Medium-risk and No-injury.Here is what we find out:
Old: Shows the highest proportion of High Risk (purple) players, with the lowest No Injury cases
Young: shows the largest proportion of No injuries and the smallest of high risk. This is quite obvious that the chances of getting injured increases as a player plays more games or ages and etc.

PCA Variable Contributions
This shows who the variables contributes to the principal components. We only used PC1 and 2 for easier visualization. And these are some variables that contributes the most

PTS (Points) 
rebounding_rate AST (Assists) 
REB (Rebounds)
PLAYER_WEIGHT DIST_MILES(distance running on court) 
PLAYER_HEIGHT_INCHES
DREB_PCT(percentage of getting defensive rebound)
OREB_PCT(percentage of getting defensive rebound)
AST_PCT(assist percentage)
USG_PCT(usage percentage)

PC Spree


```{r}
# === 3. OPTIMAL NUMBER OF CLUSTERS ===
cat("\n=== Determining Optimal Number of Clusters ===\n")

set.seed(123)

# Elbow method
elbow_plot <- fviz_nbclust(scaled_features, kmeans, method = "wss", k.max = 10) +
  labs(title = "Elbow Method for Optimal k") +
  theme_minimal()
print(elbow_plot)

# Silhouette method
silhouette_plot <- fviz_nbclust(scaled_features, kmeans, method = "silhouette", k.max = 10) +
  labs(title = "Silhouette Method for Optimal k") +
  theme_minimal()
print(silhouette_plot)

# Gap statistic (on sample due to computational cost)
set.seed(123)
if(nrow(scaled_features) > 1000) {
  sample_idx <- sample(nrow(scaled_features), 1000)
  gap_stat <- clusGap(scaled_features[sample_idx,], FUN = kmeans, K.max = 10, B = 20)
} else {
  gap_stat <- clusGap(scaled_features, FUN = kmeans, K.max = 10, B = 20)
}

gap_plot <- fviz_gap_stat(gap_stat) +
  labs(title = "Gap Statistic for Optimal k") +
  theme_minimal()
print(gap_plot)

# Choose optimal 
optimal_k <- 3 # after testing other options
cat("Using k =", optimal_k, "clusters\n")

# === 4. K-MEANS CLUSTERING ===
cat("\n=== K-Means Clustering ===\n")

set.seed(123)
kmeans_result <- kmeans(scaled_features, centers = optimal_k, nstart = 25, iter.max = 100)

# Add cluster assignments
ml_complete$kmeans_cluster <- as.factor(kmeans_result$cluster)

# Visualize K-means clusters
kmeans_plot <- fviz_cluster(kmeans_result, data = scaled_features,
                            palette = "jco",
                            geom = "point",
                            ellipse.type = "convex",
                            ggtheme = theme_minimal()) +
  labs(title = "K-Means Clustering of NBA Players")
print(kmeans_plot)

# Cluster analysis
cluster_summary <- ml_complete %>%
  group_by(kmeans_cluster) %>%
  summarise(
    n_players = n(),
    injury_rate_pct = round(mean(injured) * 100, 1),
    avg_age = round(mean(AGE), 1),
    avg_weight = round(mean(PLAYER_WEIGHT), 1),
    avg_height = round(mean(PLAYER_HEIGHT_INCHES), 1),
    avg_games = round(mean(GP), 1),
    avg_minutes = round(mean(minutes_per_game), 1),
    avg_points = round(mean(PTS), 1),
    avg_rebounds = round(mean(REB), 1),
    avg_assists = round(mean(AST), 1),
    avg_total_injuries = round(mean(total_injuries), 2),
    high_risk_pct = round(mean(injury_risk == "High_Risk") * 100, 1),
    .groups = "drop"
  )

cat("=== K-Means Cluster Profiles ===\n")
print(cluster_summary)

# === 5. HIERARCHICAL CLUSTERING ===
cat("\n=== Hierarchical Clustering ===\n")

# Use sample for computational efficiency
set.seed(123)
hc_sample_size <- min(500, nrow(scaled_features))
hc_sample_idx <- sample(nrow(scaled_features), hc_sample_size)
hc_sample_data <- scaled_features[hc_sample_idx, ]
hc_sample_ml <- ml_complete[hc_sample_idx, ]

# Distance matrix and hierarchical clustering
dist_matrix <- dist(hc_sample_data, method = "euclidean")
hc_result <- hclust(dist_matrix, method = "ward.D2")

# Dendrogram
dendro_plot <- fviz_dend(hc_result, k = optimal_k, 
                         color_labels_by_k = TRUE, rect = TRUE,
                         rect_fill = TRUE, rect_border = "jco",
                         labels_track_height = 0.1) +
  labs(title = paste("Hierarchical Clustering Dendrogram (n =", hc_sample_size, ")"))
print(dendro_plot)

# Cut tree and visualize
hc_clusters <- cutree(hc_result, k = optimal_k)
hc_sample_ml$hc_cluster <- as.factor(hc_clusters)

hc_cluster_plot <- fviz_cluster(list(data = hc_sample_data, cluster = hc_clusters),
                                palette = "jco",
                                geom = "point",
                                ellipse.type = "convex",
                                ggtheme = theme_minimal()) +
  labs(title = "Hierarchical Clustering Visualization")
print(hc_cluster_plot)

# === 6. DBSCAN CLUSTERING ===
cat("\n=== DBSCAN Clustering ===\n")

# kNN distance plot to determine eps
knn_plot <- dbscan::kNNdistplot(scaled_features, k = 4)
abline(h = 2.5, lty = 2, col = "red")

# Perform DBSCAN
set.seed(123)
dbscan_result <- dbscan::dbscan(scaled_features, eps = 2.5, minPts = 5)

ml_complete$dbscan_cluster <- as.factor(dbscan_result$cluster)

cat("DBSCAN Results:\n")
cat("Number of clusters:", max(dbscan_result$cluster), "\n")
cat("Number of noise points:", sum(dbscan_result$cluster == 0), "\n")
cat("Cluster sizes:", table(dbscan_result$cluster), "\n")

# Visualize DBSCAN
dbscan_plot <- fviz_cluster(dbscan_result, data = scaled_features,
                            palette = "jco",
                            geom = "point",
                            ellipse = FALSE,
                            ggtheme = theme_minimal()) +
  labs(title = "DBSCAN Clustering Results")
print(dbscan_plot)

# === 7. UMAP DIMENSIONALITY REDUCTION ===
cat("\n=== UMAP Analysis ===\n")

set.seed(123)
umap_result <- umap(scaled_features, n_neighbors = 15, min_dist = 0.1)

# Create UMAP data frame
umap_df <- data.frame(
  UMAP1 = umap_result$layout[,1],
  UMAP2 = umap_result$layout[,2],
  injury_risk = ml_complete$injury_risk,
  age_group = ml_complete$age_group,
  kmeans_cluster = ml_complete$kmeans_cluster,
  injured = ml_complete$injured
)

# UMAP visualization by injury risk
umap_injury_plot <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = injury_risk)) +
  geom_point(alpha = 0.7, size = 1.5) +
  scale_color_viridis_d() +
  labs(title = "UMAP - NBA Players by Injury Risk",
       x = "UMAP 1", y = "UMAP 2", color = "Injury Risk") +
  theme_minimal()
print(umap_injury_plot)

# UMAP by age group
umap_age_plot <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = age_group)) +
  geom_point(alpha = 0.7, size = 1.5) +
  scale_color_brewer(type = "qual", palette = "Set1") +
  labs(title = "UMAP - NBA Players by Age Group",
       x = "UMAP 1", y = "UMAP 2", color = "Age Group") +
  theme_minimal()
print(umap_age_plot)

# UMAP by clusters
umap_cluster_plot <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = kmeans_cluster)) +
  geom_point(alpha = 0.7, size = 1.5) +
  scale_color_brewer(type = "qual", palette = "Set1") +
  labs(title = "UMAP - NBA Players by K-Means Cluster",
       x = "UMAP 1", y = "UMAP 2", color = "Cluster") +
  theme_minimal()
print(umap_cluster_plot)

# === 8. CORRELATION ANALYSIS ===
cat("\n=== Correlation Analysis ===\n")

# Calculate correlation matrix
cor_matrix <- cor(numeric_features, use = "complete.obs")

# Correlation plot
corrplot(cor_matrix, method = "color", type = "upper", 
         order = "hclust", tl.cex = 0.7, tl.col = "black",
         title = "Feature Correlation Matrix")

# Detailed heatmap
pheatmap(cor_matrix,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         main = "Feature Correlation Heatmap",
         fontsize = 8)

# === 9. ANOMALY DETECTION ===
cat("\n=== Anomaly Detection ===\n")

# Local Outlier Factor
lof_scores <- lof(scaled_features, minPts = 5)
ml_complete$lof_score <- lof_scores
ml_complete$is_outlier <- lof_scores > quantile(lof_scores, 0.95)

cat("Outliers detected:", sum(ml_complete$is_outlier), "\n")
cat("Outlier rate:", round(mean(ml_complete$is_outlier) * 100, 2), "%\n")

# Visualize outliers
outlier_plot1 <- ggplot(ml_complete, aes(x = PTS, y = total_injuries, color = is_outlier)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("FALSE" = "blue", "TRUE" = "red")) +
  labs(title = "Outlier Detection - Points vs Injuries",
       x = "Points per Game", y = "Total Injuries", color = "Outlier") +
  theme_minimal()
print(outlier_plot1)

outlier_plot2 <- ggplot(ml_complete, aes(x = AGE, y = minutes_per_game, color = is_outlier)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("FALSE" = "blue", "TRUE" = "red")) +
  labs(title = "Outlier Detection - Age vs Minutes per Game",
       x = "Age", y = "Minutes per Game", color = "Outlier") +
  theme_minimal()
print(outlier_plot2)

# Top outliers
top_outliers <- ml_complete %>%
  filter(is_outlier) %>%
  select(PLAYER_NAME, SEASON, AGE, PTS, REB, AST, total_injuries, injury_risk, lof_score) %>%
  arrange(desc(lof_score)) %>%
  head(10)

cat("=== Top 10 Outliers ===\n")
print(top_outliers)

# === 10. CLUSTER VALIDATION ===
cat("\n=== Cluster Validation ===\n")

# Silhouette analysis for K-means
sil_analysis <- silhouette(kmeans_result$cluster, dist(scaled_features))
avg_sil_width <- mean(sil_analysis[,3])

cat("Average silhouette width:", round(avg_sil_width, 3), "\n")

# Silhouette plot
sil_plot <- fviz_silhouette(sil_analysis) +
  labs(title = "Silhouette Analysis for K-Means Clustering")
print(sil_plot)

# Within-cluster sum of squares
cat("Within-cluster sum of squares by cluster:\n")
print(kmeans_result$withinss)
cat("Total within-cluster sum of squares:", kmeans_result$tot.withinss, "\n")
cat("Between-cluster sum of squares:", kmeans_result$betweenss, "\n")

# === 11. DETAILED CLUSTER ANALYSIS ===
cat("\n=== Detailed Cluster Analysis ===\n")

# Box plots for key variables by cluster
create_cluster_boxplot <- function(var, var_name) {
  ggplot(ml_complete, aes_string(x = "kmeans_cluster", y = var, fill = "kmeans_cluster")) +
    geom_boxplot() +
    scale_fill_viridis_d() +
    labs(title = paste(var_name, "by Cluster"), 
         x = "Cluster", y = var_name, fill = "Cluster") +
    theme_minimal() +
    theme(legend.position = "none")
}

# Create multiple box plots
p1 <- create_cluster_boxplot("total_injuries", "Total Injuries")
p2 <- create_cluster_boxplot("AGE", "Age")
p3 <- create_cluster_boxplot("PTS", "Points per Game")
p4 <- create_cluster_boxplot("minutes_per_game", "Minutes per Game")

# Arrange plots
grid.arrange(p1, p2, p3, p4, ncol = 2)

# === 12. INJURY PREDICTION INSIGHTS ===
cat("\n=== Injury Prediction Insights ===\n")

# Feature importance based on cluster centers
feature_importance <- abs(kmeans_result$centers)
rownames(feature_importance) <- paste("Cluster", 1:optimal_k)

# Create feature importance heatmap
pheatmap(feature_importance,
         scale = "column",
         main = "Feature Importance by Cluster",
         cluster_cols = FALSE,
         fontsize = 8)

# Cluster-specific injury analysis
injury_by_cluster <- ml_complete %>%
  group_by(kmeans_cluster, injury_risk) %>%
  count() %>%
  group_by(kmeans_cluster) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

injury_cluster_plot <- ggplot(injury_by_cluster, aes(x = kmeans_cluster, y = prop, fill = injury_risk)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_d() +
  labs(title = "Injury Risk Distribution by Cluster",
       x = "Cluster", y = "Proportion", fill = "Injury Risk") +
  theme_minimal()
print(injury_cluster_plot)

# === 13. SAVE RESULTS ===
cat("\n=== Saving Results ===\n")

# Prepare final results dataset
final_results <- ml_complete %>%
  select(
    PLAYER_NAME, SEASON, AGE, PLAYER_WEIGHT, PLAYER_HEIGHT_INCHES,
    GP, PTS, REB, AST, total_injuries, injured, injury_risk,
    kmeans_cluster, dbscan_cluster, lof_score, is_outlier
  )

# Save results
#write.csv(final_results, "nba_unsupervised_ml_results.csv", row.names = FALSE)
#write.csv(cluster_summary, "nba_cluster_profiles.csv", row.names = FALSE)

# Save model objects for future use
#save(kmeans_result, pca_result, umap_result, scaled_features, 
     #file = "nba_ml_models.RData")

cat("Analysis complete!\n")


# === 14. SUMMARY INSIGHTS ===
cat("\n=== KEY INSIGHTS SUMMARY ===\n")
cat("1. Identified", optimal_k, "distinct player clusters\n")
cat("2. Overall injury rate:", round(mean(ml_complete$injured) * 100, 1), "%\n")
cat("3. Average silhouette width:", round(avg_sil_width, 3), "(good clustering if > 0.5)\n")
cat("4. Outliers detected:", sum(ml_complete$is_outlier), "players\n")
cat("5. PCA explains", round(sum(pca_summary$importance[2,1:3]) * 100, 1), "% variance in first 3 components\n")

# Cluster insights
for(i in 1:optimal_k) {
  cluster_data <- cluster_summary[cluster_summary$kmeans_cluster == i, ]
  cat(sprintf("   Cluster %d: %d players, %.1f%% injury rate, avg age %.1f\n", 
              i, cluster_data$n_players, cluster_data$injury_rate_pct, cluster_data$avg_age))
}

cat("\nAnalysis provides insights for injury prevention and player management strategies.\n")
``` 
Elbow method silhouette method gap statistic
we use these three plots to find the most optimal k for k-means

elbow method
Y-axis is the Total Within Sum of Square. This represents the sum of squared distances between each data point and the centroid of its assigned cluster.
The x-axis is the number of clusters.
It shows that the 

Silhouette Method for Optimal k




```{r}

# Models: Logistic Regression, XGBoost, Random Forest


# Load required libraries
library(tidyverse)
library(caret)           # For model training and evaluation
library(xgboost)         # For XGBoost
library(randomForest)    # For Random Forest
library(glmnet)          # For regularized logistic regression
library(pROC)            # For ROC curves
library(VIM)             # For missing value analysis
library(corrplot)        # For correlation plots
library(ROSE)            # For handling class imbalance
library(MLmetrics)       # For additional metrics
library(plotly)          # For interactive plots
library(lime)            # For model interpretability
library(DALEX)           # For model explanation
library(gridExtra)       # For plot arrangements
library(e1071)           # For SVM (optional)
library(ROCR)            # For performance evaluation
```
```{r}
# === DATA PREPARATION FOR SUPERVISED LEARNING ===

# Prepare the dataset
supervised_data <- merged %>%
  # Keep only rows with essential data
  filter(!is.na(PLAYER_NAME), !is.na(SEASON), !is.na(AGE)) %>%
  # Create target variable (binary injury indicator)
  mutate(
    # Target variable: injured this season (1 = injured, 0 = not injured)
    injury_target = as.factor(ifelse(injured == TRUE, "Injured", "Not_Injured")),
    
    # Feature engineering
    age_squared = AGE^2,
    weight_height_ratio = PLAYER_WEIGHT / PLAYER_HEIGHT_INCHES,
    BMI = (PLAYER_WEIGHT * 703) / (PLAYER_HEIGHT_INCHES^2),
    
    # Performance metrics per game
    minutes_per_game = ifelse(GP > 0, MIN / GP, 0),
    points_per_game = ifelse(GP > 0, PTS / GP, 0),
    rebounds_per_game = ifelse(GP > 0, REB / GP, 0),
    assists_per_game = ifelse(GP > 0, AST / GP, 0),
    
    # Efficiency metrics
    points_per_minute = ifelse(MIN > 0, PTS / MIN, 0),
    usage_efficiency = ifelse(!is.na(USG_PCT) & USG_PCT > 0, TS_PCT / USG_PCT, NA),
    rebounding_total = OREB_PCT + DREB_PCT,
    
    # Movement efficiency (if available)
    distance_per_minute = ifelse(!is.na(DIST_MILES) & MIN > 0, DIST_MILES / MIN, NA),
    speed_consistency = ifelse(!is.na(AVG_SPEED_OFF) & !is.na(AVG_SPEED_DEF), 
                              abs(AVG_SPEED_OFF - AVG_SPEED_DEF), NA),
    
    # Historical context
    season_numeric = as.numeric(SEASON),
    career_stage = case_when(
      AGE < 23 ~ "Rookie_Sophomore",
      AGE < 27 ~ "Prime_Early",
      AGE < 31 ~ "Prime_Late", 
      TRUE ~ "Veteran"
    ),
    
    # Position proxy (based on height and weight)
    position_proxy = case_when(
      PLAYER_HEIGHT_INCHES < 75 ~ "Guard",
      PLAYER_HEIGHT_INCHES < 79 ~ "Forward",
      TRUE ~ "Center"
    )
  ) %>%
  # Remove infinite values
  mutate(across(where(is.numeric), ~ifelse(is.infinite(.), NA, .)))

# Select features for modeling
feature_columns <- c(
  "AGE", "age_squared", "PLAYER_WEIGHT", "PLAYER_HEIGHT_INCHES", 
  "BMI", "weight_height_ratio",
  "GP", "minutes_per_game", "points_per_game", "rebounds_per_game", "assists_per_game",
  "points_per_minute", "PTS", "REB", "AST",
  "NET_RATING", "OREB_PCT", "DREB_PCT", "USG_PCT", "TS_PCT", "AST_PCT",
  "rebounding_total", "season_numeric"
)

# Add movement features if available
if(sum(!is.na(supervised_data$DIST_MILES)) > 100) {
  movement_features <- c("DIST_MILES", "AVG_SPEED", "distance_per_minute")
  available_movement <- movement_features[movement_features %in% names(supervised_data)]
  feature_columns <- c(feature_columns, available_movement)
  cat("Added movement features:", paste(available_movement, collapse = ", "), "\n")
}

# Create modeling dataset
modeling_data <- supervised_data %>%
  select(all_of(c("PLAYER_NAME", "SEASON", "injury_target", "career_stage", "position_proxy", feature_columns))) %>%
  # Handle missing values
  filter(complete.cases(.[feature_columns]))

cat("=== Data Summary ===\n")
cat("Total observations:", nrow(modeling_data), "\n")
cat("Features for modeling:", length(feature_columns), "\n")
cat("Class distribution:\n")
print(table(modeling_data$injury_target))
cat("Injury rate:", round(mean(modeling_data$injury_target == "Injured") * 100, 2), "%\n")

# === FEATURE ANALYSIS ===
cat("\n=== Feature Analysis ===\n")

# Correlation analysis
numeric_features <- modeling_data[feature_columns]
cor_matrix <- cor(numeric_features, use = "complete.obs")

# Plot correlation matrix
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust", 
         tl.cex = 0.7, tl.col = "black", title = "Feature Correlation Matrix")

# Feature importance via univariate analysis
univariate_analysis <- numeric_features %>%
  bind_cols(injury = modeling_data$injury_target) %>%
  pivot_longer(-injury, names_to = "feature", values_to = "value") %>%
  group_by(feature) %>%
  summarise(
    t_test_p = t.test(value ~ injury)$p.value,
    effect_size = abs(mean(value[injury == "Injured"]) - mean(value[injury == "Not_Injured"])) / sd(value),
    .groups = "drop"
  ) %>%
  arrange(t_test_p)

cat("Top 10 features by univariate significance:\n")
print(head(univariate_analysis, 10))

# === DATA SPLITTING ===
cat("\n=== Data Splitting ===\n")

set.seed(123)

# Create training and testing sets (70-30 split)
train_index <- createDataPartition(modeling_data$injury_target, p = 0.7, list = FALSE)
train_data <- modeling_data[train_index, ]
test_data <- modeling_data[-train_index, ]

cat("Training set size:", nrow(train_data), "\n")
cat("Test set size:", nrow(test_data), "\n")
cat("Training set injury rate:", round(mean(train_data$injury_target == "Injured") * 100, 2), "%\n")
cat("Test set injury rate:", round(mean(test_data$injury_target == "Injured") * 100, 2), "%\n")

# Prepare feature matrices
X_train <- as.matrix(train_data[feature_columns])
y_train <- train_data$injury_target
X_test <- as.matrix(test_data[feature_columns])
y_test <- test_data$injury_target

# === HANDLING CLASS IMBALANCE ===
cat("\n=== Handling Class Imbalance ===\n")

# Check class imbalance
class_counts <- table(y_train)
imbalance_ratio <- min(class_counts) / max(class_counts)
cat("Class imbalance ratio:", round(imbalance_ratio, 3), "\n")

# Apply SMOTE for balancing (if severely imbalanced)
if(imbalance_ratio < 0.3) {
  cat("Applying SMOTE to balance classes...\n")
  
  # Prepare data for SMOTE
  smote_data <- train_data[c(feature_columns, "injury_target")]
  
  # Apply SMOTE
  balanced_data <- ROSE(injury_target ~ ., data = smote_data, seed = 123)$data
  
  cat("Balanced dataset size:", nrow(balanced_data), "\n")
  cat("Balanced class distribution:\n")
  print(table(balanced_data$injury_target))
  
  # Update training data
  X_train_balanced <- as.matrix(balanced_data[feature_columns])
  y_train_balanced <- balanced_data$injury_target
} else {
  X_train_balanced <- X_train
  y_train_balanced <- y_train
}

# === MODEL 1: LOGISTIC REGRESSION ===
cat("\n=== Training Logistic Regression ===\n")

# Train logistic regression with regularization
set.seed(123)

# Prepare data for glmnet (requires numeric target)
y_train_numeric <- as.numeric(y_train_balanced == "Injured")
y_test_numeric <- as.numeric(y_test == "Injured")

# Cross-validation for lambda selection
cv_lasso <- cv.glmnet(X_train_balanced, y_train_numeric, family = "binomial", 
                      alpha = 1, nfolds = 5, type.measure = "auc")

# Train final model
logistic_model <- glmnet(X_train_balanced, y_train_numeric, family = "binomial", 
                        alpha = 1, lambda = cv_lasso$lambda.1se)

# Predictions
logistic_prob <- predict(logistic_model, X_test, type = "response", s = cv_lasso$lambda.1se)
logistic_pred <- ifelse(logistic_prob > 0.5, "Injured", "Not_Injured")
logistic_pred <- factor(logistic_pred, levels = c("Not_Injured", "Injured"))

# Performance metrics
logistic_cm <- confusionMatrix(logistic_pred, y_test, positive = "Injured")
logistic_auc <- roc(y_test_numeric, as.numeric(logistic_prob))$auc

cat("Logistic Regression Results:\n")
print(logistic_cm)
cat("AUC:", round(logistic_auc, 4), "\n")

# Feature importance
logistic_coef <- coef(logistic_model, s = cv_lasso$lambda.1se)
logistic_importance <- data.frame(
  feature = rownames(logistic_coef)[-1],  # Exclude intercept
  coefficient = as.numeric(logistic_coef[-1]),
  abs_coefficient = abs(as.numeric(logistic_coef[-1]))
) %>%
  arrange(desc(abs_coefficient)) %>%
  head(15)

cat("\nTop 15 Logistic Regression Features:\n")
print(logistic_importance)

# === MODEL 2: RANDOM FOREST ===
cat("\n=== Training Random Forest ===\n")

set.seed(123)

# Convert to factor for randomForest
train_rf_data <- data.frame(X_train_balanced)
train_rf_data$injury_target <- y_train_balanced

test_rf_data <- data.frame(X_test)
test_rf_data$injury_target <- y_test

# Tune Random Forest parameters
rf_grid <- expand.grid(
  mtry = c(sqrt(ncol(X_train_balanced)), ncol(X_train_balanced)/3, ncol(X_train_balanced)/2)
)

# Use caret for tuning
rf_control <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  verboseIter = FALSE
)

# Train Random Forest
rf_model <- train(
  injury_target ~ .,
  data = train_rf_data,
  method = "rf",
  tuneGrid = rf_grid,
  trControl = rf_control,
  metric = "ROC",
  ntree = 500,
  importance = TRUE
)

# Predictions
rf_prob <- predict(rf_model, test_rf_data, type = "prob")[,"Injured"]
rf_pred <- predict(rf_model, test_rf_data)

# Performance metrics
rf_cm <- confusionMatrix(rf_pred, y_test, positive = "Injured")
rf_auc <- roc(y_test_numeric, rf_prob)$auc

cat("Random Forest Results:\n")
print(rf_cm)
cat("AUC:", round(rf_auc, 4), "\n")
cat("Best mtry:", rf_model$bestTune$mtry, "\n")

# Feature importance
rf_importance <- varImp(rf_model)$importance %>%
  rownames_to_column("feature") %>%
  arrange(desc(Injured)) %>%
  head(15)

cat("\nTop 15 Random Forest Features:\n")
print(rf_importance)

# === MODEL 3: XGBOOST ===
cat("\n=== Training XGBoost ===\n")

set.seed(123)

# Prepare data for XGBoost
dtrain <- xgb.DMatrix(data = X_train_balanced, label = y_train_numeric)
dtest <- xgb.DMatrix(data = X_test, label = y_test_numeric)

# XGBoost parameters
xgb_params <- list(
  objective = "binary:logistic",
  eval_metric = "auc",
  eta = 0.1,
  max_depth = 6,
  min_child_weight = 1,
  subsample = 0.8,
  colsample_bytree = 0.8,
  seed = 123
)

# Cross-validation to find optimal rounds
xgb_cv <- xgb.cv(
  params = xgb_params,
  data = dtrain,
  nrounds = 1000,
  nfold = 5,
  showsd = TRUE,
  stratified = TRUE,
  print_every_n = 100,
  early_stopping_rounds = 50,
  maximize = TRUE
)

# Train final model
best_nrounds <- xgb_cv$best_iteration
xgb_model <- xgb.train(
  params = xgb_params,
  data = dtrain,
  nrounds = best_nrounds,
  watchlist = list(train = dtrain, test = dtest),
  print_every_n = 100,
  early_stopping_rounds = 50,
  maximize = TRUE
)

# Predictions
xgb_prob <- predict(xgb_model, dtest)
xgb_pred <- ifelse(xgb_prob > 0.5, "Injured", "Not_Injured")
xgb_pred <- factor(xgb_pred, levels = c("Not_Injured", "Injured"))

# Performance metrics
xgb_cm <- confusionMatrix(xgb_pred, y_test, positive = "Injured")
xgb_auc <- roc(y_test_numeric, xgb_prob)$auc

cat("XGBoost Results:\n")
print(xgb_cm)
cat("AUC:", round(xgb_auc, 4), "\n")
cat("Best rounds:", best_nrounds, "\n")

# Feature importance
xgb_importance <- xgb.importance(
  feature_names = colnames(X_train_balanced),
  model = xgb_model
) %>%
  head(15)

cat("\nTop 15 XGBoost Features:\n")
print(xgb_importance)

# === MODEL COMPARISON ===
cat("\n=== Model Comparison ===\n")
# === FIXED MODEL COMPARISON ===
cat("\n=== Model Comparison (Fixed) ===\n")

# Compile results with proper handling of character and numeric columns
model_results <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "XGBoost"),
  Accuracy = c(
    as.numeric(logistic_cm$overall["Accuracy"]),
    as.numeric(rf_cm$overall["Accuracy"]),
    as.numeric(xgb_cm$overall["Accuracy"])
  ),
  Sensitivity = c(
    as.numeric(logistic_cm$byClass["Sensitivity"]),
    as.numeric(rf_cm$byClass["Sensitivity"]),
    as.numeric(xgb_cm$byClass["Sensitivity"])
  ),
  Specificity = c(
    as.numeric(logistic_cm$byClass["Specificity"]),
    as.numeric(rf_cm$byClass["Specificity"]),
    as.numeric(xgb_cm$byClass["Specificity"])
  ),
  Precision = c(
    as.numeric(logistic_cm$byClass["Pos Pred Value"]),
    as.numeric(rf_cm$byClass["Pos Pred Value"]),
    as.numeric(xgb_cm$byClass["Pos Pred Value"])
  ),
  F1_Score = c(
    as.numeric(logistic_cm$byClass["F1"]),
    as.numeric(rf_cm$byClass["F1"]),
    as.numeric(xgb_cm$byClass["F1"])
  ),
  AUC = c(
    as.numeric(logistic_auc), 
    as.numeric(rf_auc), 
    as.numeric(xgb_auc)
  ),
  stringsAsFactors = FALSE
)

# Handle NA values
model_results[is.na(model_results)] <- 0

# Print results with proper formatting
cat("=== Model Performance Comparison ===\n")
cat(sprintf("%-20s %8s %11s %11s %9s %8s %6s\n", 
            "Model", "Accuracy", "Sensitivity", "Specificity", "Precision", "F1_Score", "AUC"))
cat(paste(rep("-", 80), collapse = ""), "\n")

for(i in 1:nrow(model_results)) {
  cat(sprintf("%-20s %8.4f %11.4f %11.4f %9.4f %8.4f %6.4f\n",
              model_results$Model[i],
              model_results$Accuracy[i],
              model_results$Sensitivity[i],
              model_results$Specificity[i],
              model_results$Precision[i],
              model_results$F1_Score[i],
              model_results$AUC[i]))
}

# Alternative: Create separate display for numeric results
numeric_results <- model_results[, -1]  # Exclude Model column
rownames(numeric_results) <- model_results$Model

cat("\n=== Numeric Results Only ===\n")
print(round(numeric_results, 4))

# === ENHANCED VISUALIZATION ===
cat("\n=== Creating Performance Visualizations ===\n")

# Convert to long format for plotting
model_results_long <- model_results %>%
  pivot_longer(cols = -Model, names_to = "Metric", values_to = "Value") %>%
  mutate(Value = as.numeric(Value))

# Performance comparison plot
performance_plot <- ggplot(model_results_long, aes(x = Model, y = Value, fill = Model)) +
  geom_col() +
  facet_wrap(~Metric, scales = "free_y") +
  scale_fill_viridis_d() +
  labs(title = "Model Performance Comparison", 
       subtitle = "NBA Injury Prediction Models",
       x = "Model", y = "Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

print(performance_plot)

# AUC comparison bar plot
auc_plot <- ggplot(model_results, aes(x = reorder(Model, AUC), y = AUC, fill = Model)) +
  geom_col() +
  geom_text(aes(label = round(AUC, 3)), vjust = -0.3) +
  scale_fill_viridis_d() +
  labs(title = "Model AUC Comparison", 
       subtitle = "NBA Injury Prediction",
       x = "Model", y = "AUC Score") +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(0, 1)

print(auc_plot)

# === ROBUST ERROR HANDLING FOR ROC CURVES ===
cat("\n=== ROC Curve Analysis (Enhanced) ===\n")

# Create ROC objects with error handling
tryCatch({
  roc_logistic <- roc(as.numeric(y_test == "Injured"), as.numeric(logistic_prob))
  roc_rf <- roc(as.numeric(y_test == "Injured"), rf_prob)
  roc_xgb <- roc(as.numeric(y_test == "Injured"), xgb_prob)
  
  # Enhanced ROC plot using ggplot
  roc_data <- data.frame(
    FPR = c(1 - roc_logistic$specificities, 1 - roc_rf$specificities, 1 - roc_xgb$specificities),
    TPR = c(roc_logistic$sensitivities, roc_rf$sensitivities, roc_xgb$sensitivities),
    Model = c(rep("Logistic Regression", length(roc_logistic$specificities)),
              rep("Random Forest", length(roc_rf$specificities)),
              rep("XGBoost", length(roc_xgb$specificities)))
  )
  
  roc_plot <- ggplot(roc_data, aes(x = FPR, y = TPR, color = Model)) +
    geom_line(size = 1.2) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
    scale_color_viridis_d() +
    labs(title = "ROC Curve Comparison",
         subtitle = paste("AUC: Logistic =", round(roc_logistic$auc, 3),
                          ", RF =", round(roc_rf$auc, 3),
                          ", XGBoost =", round(roc_xgb$auc, 3)),
         x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)") +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  print(roc_plot)
  
}, error = function(e) {
  cat("Error creating ROC curves:", e$message, "\n")
  cat("Using basic plotting instead...\n")
  
  # Fallback to basic plot
  plot(1-roc_logistic$specificities, roc_logistic$sensitivities, 
       type = "l", col = "red", lwd = 2,
       xlab = "False Positive Rate", ylab = "True Positive Rate",
       main = "ROC Curves Comparison")
  lines(1-roc_rf$specificities, roc_rf$sensitivities, col = "blue", lwd = 2)
  lines(1-roc_xgb$specificities, roc_xgb$sensitivities, col = "green", lwd = 2)
  abline(0, 1, lty = 2, col = "gray")
  legend("bottomright", 
         legend = c(paste("Logistic (AUC =", round(roc_logistic$auc, 3), ")"),
                    paste("Random Forest (AUC =", round(roc_rf$auc, 3), ")"),
                    paste("XGBoost (AUC =", round(roc_xgb$auc, 3), ")")),
         col = c("red", "blue", "green"), lty = 1, lwd = 2)
})

# === PREDICTION ANALYSIS WITH ERROR HANDLING ===
cat("\n=== Prediction Analysis ===\n")

# Create prediction comparison with proper error handling
tryCatch({
  pred_comparison <- data.frame(
    Player = test_data$PLAYER_NAME,
    Season = test_data$SEASON,
    Actual = as.character(y_test),
    Logistic_Prob = as.numeric(logistic_prob),
    RF_Prob = as.numeric(rf_prob),
    XGB_Prob = as.numeric(xgb_prob),
    Age = test_data$AGE,
    stringsAsFactors = FALSE
  )
  
  # Add average probability
  pred_comparison$Avg_Prob <- (pred_comparison$Logistic_Prob + 
                              pred_comparison$RF_Prob + 
                              pred_comparison$XGB_Prob) / 3
  
  # High-risk predictions
  high_risk_players <- pred_comparison %>%
    filter(Avg_Prob > 0.6) %>%  # Lowered threshold for more results
    arrange(desc(Avg_Prob))
  
  cat("High-risk players identified (avg probability > 0.6):\n")
  if(nrow(high_risk_players) > 0) {
    print(head(high_risk_players[, c("Player", "Season", "Age", "Actual", "Avg_Prob")], 10))
  } else {
    cat("No players found with average probability > 0.6\n")
    cat("Top 10 highest risk players:\n")
    top_risk <- pred_comparison %>%
      arrange(desc(Avg_Prob)) %>%
      head(10)
    print(top_risk[, c("Player", "Season", "Age", "Actual", "Avg_Prob")])
  }
  
  # Prediction distribution plot
  prob_dist_plot <- ggplot(pred_comparison, aes(x = Avg_Prob, fill = Actual)) +
    geom_histogram(alpha = 0.7, bins = 30, position = "identity") +
    scale_fill_viridis_d() +
    labs(title = "Distribution of Injury Probabilities",
         subtitle = "Average across all three models",
         x = "Average Predicted Probability", y = "Count") +
    theme_minimal()
  
  print(prob_dist_plot)
  
}, error = function(e) {
  cat("Error in prediction analysis:", e$message, "\n")
})

# === MODEL SUMMARY WITH RECOMMENDATIONS ===
cat("\n=== FINAL MODEL SUMMARY ===\n")

# Find best model
best_model_idx <- which.max(model_results$AUC)
best_model <- model_results$Model[best_model_idx]
best_auc <- model_results$AUC[best_model_idx]


cat("NBA INJURY PREDICTION MODEL RESULTS\n")

cat("\nBEST PERFORMING MODEL:", best_model, "\n")
cat("AUC Score:", round(best_auc, 4), "\n")
cat("Total features used:", length(feature_columns), "\n")
cat("Training samples:", nrow(train_data), "\n")
cat("Test samples:", nrow(test_data), "\n")

if(exists("high_risk_players") && nrow(high_risk_players) > 0) {
  cat("High-risk players identified:", nrow(high_risk_players), "\n")
}


# === SAVE RESULTS (FIXED) ===
cat("\n=== Saving Results ===\n")

# Save with error handling
#tryCatch({
  #write.csv(model_results, "nba_supervised_model_performance.csv", row.names = FALSE)
  #cat("✓ Model performance saved\n")
#}, error = function(e) cat("✗ Error saving model performance:", e$message, "\n"))

#tryCatch({
  #if(exists("pred_comparison")) {
   # write.csv(pred_comparison, "nba_injury_predictions.csv", row.names = FALSE)
    #cat("✓ Predictions saved\n")
  #}
#}, error = function(e) cat("✗ Error saving predictions:", e$message, "\n"))

#tryCatch({
#  if(exists("high_risk_players") && nrow(high_risk_players) > 0) {
#    #write.csv(high_risk_players, "nba_high_risk_players.csv", row.names = FALSE)
#    cat("✓ High-risk players saved\n")
# }
#}, error = function(e) cat("✗ Error saving high-risk players:", e$message, "\n"))
```

# Save model performance
write.csv(model_results, "nba_supervised_model_performance.csv", row.names = FALSE)
write.csv(cv_summary, "nba_supervised_cv_results.csv", row.names = FALSE)

# Save predictions
write.csv(pred_comparison, "nba_injury_predictions.csv", row.names = FALSE)
write.csv(high_risk_players, "nba_high_risk_players.csv", row.names = FALSE)

# Save feature importance
write.csv(xgb_importance, "nba_xgboost_feature_importance.csv", row.names = FALSE)
write.csv(rf_importance, "nba_rf_feature_importance.csv", row.names = FALSE)
write.csv(logistic_importance, "nba_logistic_feature_importance.csv", row.names = FALSE)

# Save models
save(logistic_model, rf_model, xgb_model, cv_lasso, 
     file = "nba_supervised_models.RData")

cat("Analysis complete!\n")
cat("Files saved:\n")
cat("- nba_supervised_model_performance.csv\n")
cat("- nba_supervised_cv_results.csv\n")
cat("- nba_injury_predictions.csv\n")
cat("- nba_high_risk_players.csv\n")
cat("- Feature importance files (3 models)\n")
cat("- nba_supervised_models.RData\n")



